---
title: "R Notebook: exploration of various timers"
output: html_notebook
---

Read in the output of `go run ./src/instrumentation_measurement/clock_resolution_go/main.go`

```{r fig.width=20}
setwd("~/sources/imapp/gas-cost-estimator/src")
time_all = read.csv("../../time.csv")
N = 50000
time = head(time_all, N)
head(time)
```
The temporal dynamics of all timers must be accounted for. They all seem to warm up for a long time.

**NOTE** `purple` line (gotsc) is not in ns but in CPU cycles

```{r fig.width=20}
plot(NULL, xlim=c(1, N), ylim=c(0, 3000))
lines(time$clock_gettime, type = "l", col = "red")
lines(time$time, type = "l", col = "blue")
lines(time$runtime_nano, type = "l", col = "green")
lines(time$gotsc, type = "l", col = "purple")
```

```{r fig.width=20}
plot(NULL, xlim=c(1, N), ylim=c(1, 3.5))
par(ylog=TRUE)
lines(time$clock_gettime, type = "l", col = "red")
lines(time$time, type = "l", col = "blue")
lines(time$runtime_nano, type = "l", col = "green")
lines(time$gotsc, type = "l", col = "purple")
```

It seems `runtimeNano` is the most accurate and stable one. We could perhaps subtract the `Min.` of this from all the measurements

```{r fig.width=20}
summary(time)
```

```{r fig.width=20}
boxplot(time)
```

Explore the effect of the overhead increasing for all timers. We're trimming down the data frame to observe correlations sensibly:
```{r fig.width=20}
time_sample = time[sample(nrow(time), 100), ]
var(time_sample)
cor(time_sample)

# cleanup
rm(time_sample)
```
Deeper analysis of the two best clocs: `runtimeNano` and `gotsc`:

```{r fig.width=20}
par(mfrow=c(2,1))
frequencies = sort(table(time$runtime_nano), decreasing=TRUE)
plot(frequencies[1:50], col="green")
frequencies = sort(table(time$gotsc), decreasing=TRUE)
plot(frequencies[1:50], col="purple")

# cleanup
rm(frequencies)
```

```{r fig.width=20}
quantile(time$runtime_nano, probs=c(0.85, 0.9, 0.95, 0.99, 0.999, 0.9999, 0.99999))
quantile(time$gotsc, probs=c(0.85, 0.9, 0.95, 0.99, 0.999, 0.9999, 0.99999))
```

Summary:

1. We should discard about 5000 first observations (**TODO**)
2. It is probably a good idea to monitor and register the timer overhead during the opcode measurements (**TODO**).
2. Due to periods of increased overhead, sometimes the measurements might be over-timed for several consecutive measurements. Should we discard all measurements where "just time" measurement is above a threshold? (**TODO**)
3. `runtimeNano` is clearly the winner, but it still has high values quite often, and is subject to large overhead during warm-up and during the "temporary increase periods"
    - **UPDATE** - it is a winner in wall-clock category, but probably CPU cycles using `gotsc` (based on TSC and in-sync with what `evmone` measurements use - RDTSC) is even better
4. We can also subtract the minimum (or mean/median) observed timer overhead of `runtimeNano` from all the measurements
    - **UPDATE** - if we go for CPU cycles it's 35. Interestingly though, the `gotsc` library tells us "TSC Overhead: 31"
5. Next step would be to consider subtracting more, considering it's a justified move (**TODO**)
6. Alternatively, we could do more in-depth analysis of the behaviors observed and try to normalize the timer readings more (**TODO** optional)

**TODO** also another timer, another one tried by `chfast` for evmone https://godoc.org/github.com/lanl/go-papi
**TODO** explore C++ and Rust timers similarly
