---
title: "Gas Cost Estimator: comparison of obtained estimates, alternative gas cost schedule"
output: html_document
params:
  estimate_files: ""
  current_gas_cost: ""
---

```{r, setup, include=FALSE}
#  params$estimate_files = "besu_marginal_estimated_cost.csv, erigon_marginal_estimated_cost.csv, ethereumjs_marginal_estimated_cost.csv, geth_marginal_estimated_cost.csv, nethermind_marginal_estimated_cost.csv, revm_marginal_estimated_cost.csv"
#  params$current_gas_cost = "current_gas_cost.csv"

if (params$estimate_files == "") {
  stop("'estimate_files' param is missing, the comma separated list of estimate files")
}

if (params$current_gas_cost == "") {
  stop("'current_gas_cost' param is missing, the file with current gas cost schedule")
}

library(sqldf)
library(nlme)
library(mixtools)
library(zeallot) # for multi-assignment %<-%

knitr::opts_knit$set(root.dir = '/data')

geth_color = rgb(0.1,0.1,0.7,0.5)
evmone_color = rgb(0.8,0.1,0.3,0.6)
besu_color = rgb(0.5,0.1,0.5,0.4)
nethermind_color = rgb(0.4,0.2,0.6,0.8)
ethereumjs_color = rgb(0.3,0.8,0.2,0.6)
erigon_color = rgb(0.8,0.8,0.1,0.7)
revm_color = rgb(0.2,0.7,0.8,0.3)

env_colors = c(geth_color, evmone_color, besu_color, nethermind_color, ethereumjs_color, erigon_color, revm_color)
```

In this script we conduct the comparison procedure of the obtained estimates.

```{r, include=FALSE}
all_estimate_files = strsplit(params$estimate_files, split="\\s*,\\s*")[[1]]

if (length(all_estimate_files) == 0) {
  stop("No estimated cost file found.")
}
# TODO parametrization
# TODO details
estimated_cost = data.frame(op=character(0),estimate_marginal_ns=numeric(0),estimate_marginal_stderr=numeric(0),env=character(0))
for (i in 1:length(all_estimate_files)) {
  add_estimated_cost = read.csv(all_estimate_files[i])
  estimated_cost = rbind(estimated_cost, add_estimated_cost[, c("op", "estimate_marginal_ns", "estimate_marginal_ns_stderr", "env")])
}
if (nrow(estimated_cost) == 0) {
  stop("No estimated cost data found.")
}

all_opcodes = unique(estimated_cost$op)
all_envs = unique(estimated_cost$env)

current_gas_cost = read.csv(params$current_gas_cost)
```

```{r, include=FALSE}
# df - must contain all opcodes from  reference_df
# reference_df - subset of current_gas_cost
calculate_l2_scale <- function(df, reference_df) {
  q = "SELECT avg(estimate_marginal_ns/constant_current_gas) FROM reference_df INNER JOIN df ON df.op = reference_df.opcode"
  sqldf(q)
}

calculate_l2_row_mean <- function(df) {
  out_df = data.frame(matrix(nrow = nrow(df), ncol = 0))
  out_df$mean <- rowMeans(df, na.rm = TRUE)
  mf <- function(x){sd(df[x, ])/sqrt(length(all_envs))}
  out_df$stderr <- sapply(1:nrow(df), mf)
  out_df
}

calculate_scale = calculate_l2_scale
calculate_row_mean = calculate_l2_row_mean
```

```{r, include=FALSE}
q = c("SELECT ")
for (env in all_envs) {
  df = estimated_cost[which(estimated_cost$env==env),]
  eligible_opcodes = df[,'op'] # TODO limit opcodes for scale
  reference_df = current_gas_cost[which(current_gas_cost$groups == 0),]
  scale = calculate_scale(df, reference_df)
  q <- append(q, paste0(env, ".estimate_marginal_ns/", scale, " as ", env, "_gas,"))
  q <- append(q, paste0(env, ".estimate_marginal_ns_stderr/", scale, " as ", env, "_gas_stderr,"))
}
q <- append(q, " current_gas_cost.opcode, current_gas_cost.constant_current_gas FROM current_gas_cost")
for (env in all_envs) {
  q <- append(q, paste0(" INNER JOIN estimated_cost as ", env, " ON ", env, ".op = current_gas_cost.opcode AND ", env, ".env = '", env, "'"))
}
q <- append(q, " WHERE current_gas_cost.groups >= 0")
q

gas_schedule_comparison <- sqldf(paste(q, collapse = ""))

all_envs_cols <- paste0(all_envs, "_gas")

max_groups = max(current_gas_cost[, "groups"])
```

Plot the scaled gas cost schedules.

```{r fig.width=20, fig.height=15}
for (g in 0:max_groups) {
  groups_opcode <- current_gas_cost[which(current_gas_cost$groups == g), "opcode"]
  if (length(groups_opcode) == 0) {
    next
  }
  grups_gas_schedule_comparison <- gas_schedule_comparison[which(gas_schedule_comparison$opcode %in% groups_opcode), ]
  if (nrow(grups_gas_schedule_comparison) == 0) {
    next
  }
  maximum_gas_cost = max(grups_gas_schedule_comparison[, c(all_envs_cols, "constant_current_gas")])
  plot(grups_gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.2), xlab="", ylab="", las=2)
  axis(1, at=1:nrow(grups_gas_schedule_comparison), labels=grups_gas_schedule_comparison$opcode, las=2)
  for (i in 1:length(all_envs)) {
    points(grups_gas_schedule_comparison[, paste0(all_envs[i], "_gas")], col=env_colors[i], bg=env_colors[i], xaxt='n', pch=21)
  }
  title(main="The scaled gas costs")
  legend(1, maximum_gas_cost * 1.2, c(all_envs, "Current gas cost schedule"), fill=c(env_colors[1:length(all_envs)], 'grey'), cex=2)
}
```

And individual plots of scaled gas schedules for EVMs.

```{r fig.width=20, fig.height=15}
for (i in 1:length(all_envs)) {
  for (g in 0:max_groups) {
    groups_opcode <- current_gas_cost[which(current_gas_cost$groups == g), "opcode"]
    if (length(groups_opcode) == 0) {
      next
    }
    grups_gas_schedule_comparison <- gas_schedule_comparison[which(gas_schedule_comparison$opcode %in% groups_opcode), c(paste0(all_envs[i], "_gas"), "opcode", "constant_current_gas")]
    if (nrow(grups_gas_schedule_comparison) == 0) {
      next
    }
    maximum_gas_cost = max(grups_gas_schedule_comparison[, c(paste0(all_envs[i], "_gas"), "constant_current_gas")])
    plot(grups_gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.2), xlab="", ylab="")
    axis(1, at=1:nrow(grups_gas_schedule_comparison), labels=grups_gas_schedule_comparison$opcode, las=2)
    points(grups_gas_schedule_comparison[, paste0(all_envs[i], "_gas")], col=env_colors[i], bg=env_colors[i], xaxt='n', pch=21)
    legend(1, maximum_gas_cost * 1.2, c(all_envs[i], "Current gas cost schedule"), fill=c(env_colors[i], 'grey'), cex=2)
    title(main=paste0("The scaled gas cost - ", all_envs[i]))
  }
}
```

Calculate and plot the alternative gas cost schedule.
The calculated values are:

* alternative_gas: the average of evms gas schedules, without any weights,
* alternative_gas_stderr: the standard deviation of evms gas schedules,
* alternative_gas_rel_diff: the relative difference between the current gas schedule and the alternative gas schedule, as a fraction,
* alternative_gas_rel_stderr: the relative standard deviation of evms gas schedules, relative to the alternative gas schedule, as a fraction.

```{r, include=FALSE}
if (length(all_envs_cols) == 1) { # special case
  gas_schedule_comparison$alternative_gas <- gas_schedule_comparison[, all_envs_cols]
  gas_schedule_comparison$alternative_gas_stderr <- 0
} else {
  alternative_gas = calculate_row_mean(gas_schedule_comparison[, all_envs_cols])
  gas_schedule_comparison$alternative_gas = alternative_gas$mean
  gas_schedule_comparison$alternative_gas_stderr = alternative_gas$stderr
}

gas_schedule_comparison$alternative_gas_rel_stderr <- gas_schedule_comparison$alternative_gas_stderr / gas_schedule_comparison$alternative_gas
gas_schedule_comparison$alternative_gas_rel_diff <- abs((gas_schedule_comparison$constant_current_gas - gas_schedule_comparison$alternative_gas) / gas_schedule_comparison$constant_current_gas)
```

```{r}
gas_schedule_comparison[, c('opcode', 'constant_current_gas', 'alternative_gas', 'alternative_gas_stderr', 'alternative_gas_rel_diff', 'alternative_gas_rel_stderr')]
```

```{r fig.width=20, fig.height=10}
for (g in 0:max_groups) {
  groups_opcode <- current_gas_cost[which(current_gas_cost$groups == g), "opcode"]
  if (length(groups_opcode) == 0) {
    next
  }
  grups_gas_schedule_comparison <- gas_schedule_comparison[which(gas_schedule_comparison$opcode %in% groups_opcode), c("opcode", "constant_current_gas", "alternative_gas")]
  if (nrow(grups_gas_schedule_comparison) == 0) {
    next
  }
  maximum_gas_cost = max(grups_gas_schedule_comparison[, c("constant_current_gas", "alternative_gas")])
  
  plot(grups_gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.1), xlab="", ylab="")
  axis(1, at=1:nrow(grups_gas_schedule_comparison), labels=grups_gas_schedule_comparison$opcode, las=2)
  points(grups_gas_schedule_comparison$alternative_gas, col='orange', bg='orange', xaxt='n', pch=21)
  title(main="The alternative gas cost schedule")
  legend(0, maximum_gas_cost, c("Alternative gas cost schedule", "Current gas cost schedule"), fill=c('orange', 'grey'), cex=2)
}
```

```{r fig.width=30, fig.height=15, include=FALSE}
# This is just for ad hoc visualization
maximum_gas_cost = 13
max_values = nrow(gas_schedule_comparison) - 10
constant_current_gas_color = rgb(0.4,0.4,0.4,1.0)
alternative_gas_color = rgb(0.1,0.2,1.0,1.0)

plot(gas_schedule_comparison[1:max_values, 'constant_current_gas'], col=constant_current_gas_color, bg=constant_current_gas_color, xaxt='n', pch=21, ylim=c(0,maximum_gas_cost * 1.1), xlab="", ylab="")
axis(1, at=1:max_values, labels=gas_schedule_comparison[1:max_values, 'opcode'], las=2)
points(gas_schedule_comparison[1:max_values, 'alternative_gas'], col=alternative_gas_color, bg=alternative_gas_color, xaxt='n', pch=21, cex=1.5)
legend(75, maximum_gas_cost, c("Alternative gas cost schedule", "Current gas cost schedule"), fill=c(alternative_gas_color, constant_current_gas_color), cex=2)
```

```{r, include=FALSE}
# just export the final compiled estimations to a file
# write.csv(estimated_cost, "final_estimated_cost.csv", quote=FALSE, row.names=FALSE) # estimated_cost is changed in stage4
# write.csv(gas_schedule_comparison, "gas_schedule_comparison.csv", quote=FALSE, row.names=FALSE)
```