---
title: "R Notebook: comparison of obtained estimates for all envs - the final results; for Gas Cost Estimator"
output: html_document
---
In this script we conduct the comparison procedure of the obtained estimates.

```{r, setup, include=FALSE}
# NOTE: change this to your respective path
knitr::opts_knit$set(root.dir = '~/sources/imapp3/gas-cost-estimator/src')
source('common.R')
all_envs = c("geth", "evmone", "besu", "nethermind", "ethereumjs", "erigon", "revm")
```

```{r}
current_gas_cost = read.csv("program_generator/data/current_gas_cost.csv")

if (exists("estimated_cost")) {
  rm(estimated_cost)
}
for (env in all_envs) {
  if (exists("estimated_cost")) {
    estimated_cost = rbind(estimated_cost, read.csv(paste0("../../local/", env, "_argument_estimated_cost.csv")))
  } else {
    estimated_cost = read.csv(paste0("../../local/", env, "_argument_estimated_cost.csv"))
  }
}

if (exists("marginal")) {
  rm(marginal)
}
for (env in all_envs) {
  if (exists("marginal")) {
    marginal = rbind(marginal, read.csv(paste0("../../local/", env, "_marginal_estimated_cost.csv")))
  } else {
    marginal = read.csv(paste0("../../local/", env, "_marginal_estimated_cost.csv"))
  }
}

order1 = order(estimated_cost$opcode, estimated_cost$env)
order2 = order(marginal$op, marginal$env)

estimated_cost[order1, 'estimate_marginal_ns'] = marginal[order2, 'estimate_marginal_ns']
estimated_cost[order1, 'estimate_marginal_ns_stderr'] = marginal[order2, 'estimate_marginal_ns_stderr']

estimated_cost$arg0_ns[which(is.na(estimated_cost$arg0_ns))] = 0
estimated_cost$arg1_ns[which(is.na(estimated_cost$arg1_ns))] = 0
estimated_cost$arg2_ns[which(is.na(estimated_cost$arg2_ns))] = 0
estimated_cost$expensive_ns[which(is.na(estimated_cost$expensive_ns))] = 0

head(estimated_cost)
head(marginal)
```

Here we calculate the scales. Each estimated evm gas cost is scaled to get the best match with the current gas schedule.
There are many PUSH, DUP and SWAP opcodes.
To avoid other opcodes being overran, we use only PUSH1, DUP1 and SWAP1 as representatives when calculating the scales.

```{r}
# [SECTION] original

extract_opcodes <- function() {
  unique(estimated_cost$opcode)
}
all_opcodes = extract_opcodes()
excluded_opcodes = c('PUSH2', 'PUSH3', 'PUSH4', 'PUSH5', 'PUSH6', 'PUSH7', 'PUSH8', 'PUSH9', 'PUSH10', 'PUSH11', 'PUSH12', 'PUSH13', 'PUSH14', 'PUSH15', 'PUSH16', 
                     'PUSH17', 'PUSH18', 'PUSH19', 'PUSH20', 'PUSH21', 'PUSH22', 'PUSH23', 'PUSH24', 'PUSH25', 'PUSH26', 'PUSH27', 'PUSH28', 'PUSH29', 'PUSH30', 
                     'PUSH31', 'PUSH32', 
                     'DUP2', 'DUP3', 'DUP4', 'DUP5', 'DUP6', 'DUP7', 'DUP8', 'DUP9', 'DUP10', 'DUP11', 'DUP12', 'DUP13', 'DUP14', 'DUP15', 'DUP16', 
                     'SWAP2', 'SWAP3', 'SWAP4', 'SWAP5', 'SWAP6', 'SWAP7', 'SWAP8', 'SWAP9', 'SWAP10', 'SWAP11', 'SWAP12', 'SWAP13', 'SWAP14', 'SWAP15', 'SWAP16')

```

```{r}
excluded_opcodes_df <- data.frame(excluded_opcodes)
colnames(excluded_opcodes_df) <- 'opcode'
q = c("SELECT ")
for (env in all_envs) {
  q <- append(q, paste0("avg(", env, ".estimate_marginal_ns/current_gas_cost.constant_current_gas) as ", env, "_scale,"))
}
q <- append(q, " 1 FROM current_gas_cost")
for (env in all_envs) {
  q <- append(q, paste0(" INNER JOIN estimated_cost as ", env, " ON ", env, ".opcode = current_gas_cost.opcode AND ", env, ".env = '", env, "'"))
}
q <- append(q, paste0(" WHERE current_gas_cost.opcode NOT IN (SELECT opcode FROM excluded_opcodes_df)"))
scales <- sqldf(paste(q, collapse = ""))[1:length(all_envs)]
```

This allows us to have gas cost schedules with the least divergence between EVMs.

```{r}
q = c("SELECT ")
for (env in all_envs) {
  q <- append(q, paste0(env, ".estimate_marginal_ns/", env, "_scale as ", env, "_gas,"))
}
for (env in all_envs) {
  q <- append(q, paste0(env, ".estimate_marginal_ns_stderr/", env, "_scale as ", env, "_gas_stderr,"))
}
q <- append(q, " current_gas_cost.opcode, current_gas_cost.constant_current_gas FROM current_gas_cost, scales")
for (env in all_envs) {
  q <- append(q, paste0(" INNER JOIN estimated_cost as ", env, " ON ", env, ".opcode = current_gas_cost.opcode AND ", env, ".env = '", env, "'"))
}

gas_schedule_comparison <- sqldf(paste(q, collapse = ""))
```

We continue to fill in the additional special entries of our gas cost schedule, being the costs of the arguments.

```{r}
arg_cost <- function(opcode, arg, opcode_name, opcode_current_gas_cost) {
  r1 <- list()
  r2 <- list()
  for (env in all_envs) {
    r1 <- append(r1, estimated_cost[which(estimated_cost$env==env & estimated_cost$opcode==opcode),paste0('arg', arg, '_ns')])
    r2 <- append(r2, estimated_cost[which(estimated_cost$env==env & estimated_cost$opcode==opcode),paste0('arg', arg, '_ns_stderr')])
  }
  append(append(r1/scales, r2/scales), list(opcode_name, opcode_current_gas_cost))
}

gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = arg_cost('EXP', '1', "EXP_arg1_cost", 50)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = arg_cost('CALLDATACOPY', '2', "CALLDATACOPY_arg2_cost", 3)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = arg_cost('RETURNDATACOPY', '2', "RETURNDATACOPY_arg2_cost", 3)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = arg_cost('CODECOPY', '2', "CODECOPY_arg2_cost", 3)
```

```{r}
expensive_cost <- function(opcode, opcode_name, opcode_current_gas_cost) {
  r1 <- list()
  r2 <- list()
  for (env in all_envs) {
    v1 <- estimated_cost$expensive_ns[which(estimated_cost$env==env & estimated_cost$opcode==opcode)] + estimated_cost$estimate_marginal_ns[which(estimated_cost$env==env & estimated_cost$opcode==opcode)]
    r1 <- append(r1, v1)
    v2 <- sqrt(estimated_cost$expensive_ns_stderr[which(estimated_cost$env==env & estimated_cost$opcode==opcode)]^2 + estimated_cost$estimate_marginal_ns_stderr[which(estimated_cost$env==env & estimated_cost$opcode==opcode)]^2)
    r2 <- append(r2, v2) 
  }
  append(append(r1/scales, r2/scales), list(opcode_name, opcode_current_gas_cost))
}

gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = expensive_cost('DIV', "DIV_expensive_cost", 5)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = expensive_cost('SDIV', "SDIV_expensive_cost", 5)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = expensive_cost('MOD', "MOD_expensive_cost", 5)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = expensive_cost('SMOD', "SMOD_expensive_cost", 5)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = expensive_cost('ADDMOD', "ADDMOD_expensive_cost", 8)
gas_schedule_comparison[nrow(gas_schedule_comparison) + 1, ] = expensive_cost('MULMOD', "MULMOD_expensive_cost", 8)
```

```{r}
gas_schedule_comparison[, c(which(names(gas_schedule_comparison) == 'opcode'), 1:length(all_envs))]
```

Plot the scaled gas cost schedules.

```{r fig.width=30, fig.height=10}
maximum_gas_cost = max(gas_schedule_comparison[,!names(gas_schedule_comparison) %in% c('opcode')])
curr_env_colors = env_colors[1:length(all_envs)]

plot(gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.1), xlab="", ylab="")
axis(1, at=1:nrow(gas_schedule_comparison), labels=gas_schedule_comparison$opcode, las=2)
for (i in 1:length(all_envs)) {
  points(gas_schedule_comparison[,i], col=curr_env_colors[i], bg=curr_env_colors[i], xaxt='n', pch=21)
}
legend(0, maximum_gas_cost, c(all_envs, "Current gas cost schedule"), fill=c(curr_env_colors, 'grey'), cex=2)
```

And individual plots of scaled gas schedules for EVMs.

```{r fig.width=30, fig.height=10}
maximum_gas_cost = max(gas_schedule_comparison[,!names(gas_schedule_comparison) %in% c('opcode')])

for (i in 1:length(all_envs)) {
  plot(gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.1), xlab="", ylab="")
  axis(1, at=1:nrow(gas_schedule_comparison), labels=gas_schedule_comparison$opcode, las=2)
  points(gas_schedule_comparison[,i], col=curr_env_colors[i], bg=curr_env_colors[i], xaxt='n', pch=21)
  legend(0, maximum_gas_cost, c(all_envs[i], "Current gas cost schedule"), fill=c(env_colors[i], 'grey'), cex=2)
}
```

Calculate and plot the alternative gas cost schedule.
The calculated values are:
- alternative_gas: the average of evms gas schedules, without any weights,
- alternative_gas_stderr: the standard deviation of evms gas schedules,
- alternative_gas_rel_diff: the relative difference between the current gas schedule and the alternative gas schedule, as a fraction,
- alternative_gas_rel_stderr: the relative standard deviation of evms gas schedules, relative to the alternative gas schedule, as a fraction.

```{r}
gas_schedule_comparison$alternative_gas <- rowMeans(gas_schedule_comparison[, 1:length(all_envs)], na.rm = TRUE)
mf <- function(x){sd(gas_schedule_comparison[x,1:length(all_envs)])/sqrt(length(all_envs))}
gas_schedule_comparison$alternative_gas_stderr <- sapply(1:nrow(gas_schedule_comparison),mf)
gas_schedule_comparison$alternative_gas_rel_stderr <- gas_schedule_comparison$alternative_gas_stderr / gas_schedule_comparison$alternative_gas
gas_schedule_comparison$alternative_gas_rel_diff <- abs((gas_schedule_comparison$constant_current_gas - gas_schedule_comparison$alternative_gas) / gas_schedule_comparison$constant_current_gas)

gas_schedule_comparison[, c('opcode', 'constant_current_gas', 'alternative_gas', 'alternative_gas_rel_diff', 'alternative_gas_rel_stderr')]
```

```{r fig.width=30, fig.height=10}
maximum_gas_cost = max(gas_schedule_comparison[,!names(gas_schedule_comparison) %in% c('opcode')])

plot(gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.1), xlab="", ylab="")
axis(1, at=1:nrow(gas_schedule_comparison), labels=gas_schedule_comparison$opcode, las=2)
points(gas_schedule_comparison$alternative_gas, col='orange', bg='orange', xaxt='n', pch=21)
legend(0, maximum_gas_cost, c("Alternative gas cost schedule", "Current gas cost schedule"), fill=c('orange', 'grey'), cex=2)
```
```{r fig.width=30, fig.height=15}
maximum_gas_cost = 13
max_values = nrow(gas_schedule_comparison) - 10
constant_current_gas_color = rgb(0.4,0.4,0.4,1.0)
alternative_gas_color = rgb(0.1,0.2,1.0,1.0)

plot(gas_schedule_comparison[1:max_values, 'constant_current_gas'], col=constant_current_gas_color, bg=constant_current_gas_color, xaxt='n', pch=21, ylim=c(0,maximum_gas_cost * 1.1), xlab="", ylab="")
axis(1, at=1:max_values, labels=gas_schedule_comparison[1:max_values, 'opcode'], las=2)
points(gas_schedule_comparison[1:max_values, 'alternative_gas'], col=alternative_gas_color, bg=alternative_gas_color, xaxt='n', pch=21, cex=1.5)
legend(75, maximum_gas_cost, c("Alternative gas cost schedule", "Current gas cost schedule"), fill=c(alternative_gas_color, constant_current_gas_color), cex=2)
```

```{r}
# just export the final compiled estimations to a file
write.csv(estimated_cost, "../../local/final_estimated_cost.csv", quote=FALSE, row.names=FALSE)
write.csv(gas_schedule_comparison, "../../local/gas_schedule_comparison.csv", quote=FALSE, row.names=FALSE)
```