---
title: "Gas Cost Estimator: comparison of obtained estimates, alternative gas cost schedule"
output: html_document
params:
  estimate_files: "/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_create_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_create_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_create_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_create_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_create_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_create_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_create_revm.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_full_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_full_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_full_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_full_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_full_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_full_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_full_revm.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_mem_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_mem_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_mem_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_mem_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_mem_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_mem_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_mem_revm.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_precompiles_full_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_precompiles_full_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_precompiles_full_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_precompiles_full_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_precompiles_full_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_stop_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_stop_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_stop_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_stop_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_stop_revm.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_t_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_t_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_t_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_t_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_t_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_t_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_marginal_t_revm.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_arithmetic_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_arithmetic_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_arithmetic_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_arithmetic_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_arithmetic_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_arithmetic_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_arithmetic_revm.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_create_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_more_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_more_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_more_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_more_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_more_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_more_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_more_revm.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_precompiles_besu.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_precompiles_erigon.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_precompiles_ethereumjs.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_precompiles_evmone.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_precompiles_geth.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_precompiles_nethermind.csv,/home/lukaszglen/sources/imapp5/local/reports-12.01.2025/estimated_cost_arguments_precompiles_revm.csv"
  current_gas_cost: "/home/lukaszglen/sources/imapp5/gas-cost-estimator/src/analysis/current_gas_cost.csv"
  output_comparison_file: ""
  metric: "l1"
  details: "103"
---

```{r, setup, include=FALSE}
if (params$estimate_files == "") {
  stop("'estimate_files' param is missing, the comma separated list of estimate files")
}

if (params$current_gas_cost == "") {
  stop("'current_gas_cost' param is missing, the file with current gas cost schedule")
}

if (!(params$metric %in% c("l1", "l2"))) {
  stop("unsupported metric")
}

if (params$details == "") {
  details_arg = 0
  details = FALSE
} else {
  suppressWarnings(details_arg <- as.integer(params$details))
  if (is.na(details_arg)) {
    details_arg = 0
    details = tolower(params$details) %in% c("1", "t", "true", "on")
  } else {
    details = details_arg > 0
  }
}

library(sqldf)
library(nlme)
library(mixtools)
library(zeallot) # for multi-assignment %<-%

knitr::opts_knit$set(root.dir = '/data')

geth_color = rgb(0.1,0.1,0.7,0.5)
evmone_color = rgb(0.8,0.1,0.3,0.6)
besu_color = rgb(0.5,0.1,0.5,0.4)
nethermind_color = rgb(0.4,0.2,0.6,0.8)
ethereumjs_color = rgb(0.3,0.8,0.2,0.6)
erigon_color = rgb(0.8,0.8,0.1,0.7)
revm_color = rgb(0.2,0.7,0.8,0.3)

env_colors = c(geth_color, evmone_color, besu_color, nethermind_color, ethereumjs_color, erigon_color, revm_color)
```

In this script we conduct the comparison procedure of the obtained estim

Parametrization. 
The comma separeted list of csv files with precomputed estimated costs for evm clients estimate_files=`r params$estimate_files`,
the file with the reference gas cost table and groups configuration current_gas_cost=`r params$current_gas_cost`,
used metric code metric=`r params$metric`,
should the details be included in the report details=`r details`.

```{r, include=FALSE}
all_estimate_files = strsplit(params$estimate_files, split="\\s*,\\s*")[[1]]

if (length(all_estimate_files) == 0) {
  stop("No estimated cost file found.")
}
# TODO details
estimated_cost = data.frame(op=character(0),estimate_marginal_ns=numeric(0),estimate_marginal_ns_stderr=numeric(0),env=character(0))
for (i in 1:length(all_estimate_files)) {
  add_estimated_cost_raw = read.csv(all_estimate_files[i])
  if (all(c("opcode", "arg0_ns_raw", "arg1_ns_raw", "arg2_ns_raw", "env") %in% colnames(add_estimated_cost_raw))) {
    add_estimated_cost = data.frame(op=character(0),estimate_marginal_ns=numeric(0),estimate_marginal_ns_stderr=numeric(0),env=character(0))
    add_estimated_cost_arg0 <- sqldf("SELECT opcode || '_ARG0' AS op, arg0_ns_raw AS estimate_marginal_ns, 0 AS estimate_marginal_ns_stderr, env FROM add_estimated_cost_raw WHERE arg0_ns_raw IS NOT NULL")
    add_estimated_cost_arg1 <- sqldf("SELECT opcode || '_ARG1' AS op, arg1_ns_raw AS estimate_marginal_ns, 0 AS estimate_marginal_ns_stderr, env FROM add_estimated_cost_raw WHERE arg1_ns_raw IS NOT NULL")
    add_estimated_cost_arg2 <- sqldf("SELECT opcode || '_ARG2' AS op, arg2_ns_raw AS estimate_marginal_ns, 0 AS estimate_marginal_ns_stderr, env FROM add_estimated_cost_raw WHERE arg2_ns_raw IS NOT NULL")
    add_estimated_cost_argc <- sqldf("SELECT opcode || '_ARGC' AS op, estimate_marginal_ns + IFNULL(expensive_ns_raw, 0) AS estimate_marginal_ns, 0 AS estimate_marginal_ns_stderr, env FROM add_estimated_cost_raw WHERE estimate_marginal_ns IS NOT NULL")
    add_estimated_cost = rbind(add_estimated_cost, add_estimated_cost_arg0, add_estimated_cost_arg1, add_estimated_cost_arg2, add_estimated_cost_argc)
  } else if (all(c("op", "estimate_marginal_ns", "estimate_marginal_ns_stderr", "env") %in% colnames(add_estimated_cost_raw))) { # it is marginal
    add_estimated_cost <- add_estimated_cost_raw
  } else {
    stop(paste("Unrecognized file format: ", all_estimate_files[i]))
  }
  estimated_cost = rbind(estimated_cost, add_estimated_cost[, c("op", "estimate_marginal_ns", "estimate_marginal_ns_stderr", "env")])
}
if (nrow(estimated_cost) == 0) {
  stop("No estimated cost data found.")
}

all_opcodes = unique(estimated_cost$op)
all_envs = unique(estimated_cost$env)

current_gas_cost = read.csv(params$current_gas_cost)
```

```{r, include=FALSE}
# simple validation of uniqueness of estimates
estimated_cost_non_unique <- sqldf("SELECT op, env FROM estimated_cost GROUP BY op, env HAVING COUNT(1) > 1")
if (nrow(estimated_cost_non_unique) > 0) {
  stop(paste("Non unique input data, estimate cost multiplied for:", estimated_cost_non_unique[1, 'op'], estimated_cost_non_unique[1, 'env']))
}
```

```{r, include=FALSE}
# reference_df - subset of current_gas_cost
calculate_l1_scale <- function(df, reference_df) {
  q = "SELECT estimate_marginal_ns/constant_current_gas FROM reference_df LEFT JOIN df ON df.op = reference_df.opcode"
  values <- sqldf(q)
  m <- median(values[,1], na.rm=TRUE)
  if (is.na(m)) {
    stop("cannot calculate median")
  }
  m
}

calculate_l1_row_mean <- function(df) {
  out_df = data.frame(matrix(nrow = nrow(df), ncol = 0))
  out_df$mean <- apply(df, 1, median, na.rm=TRUE)

  mf <- function(row_) {
    m <- median(row_, na.rm = TRUE)
    if (is.na(m)) {
      NA
    } else {
      mean(sapply(row_,  function(x) ifelse(is.na(x), NA, abs(x - m)) ), na.rm=TRUE)
    }
  }
  out_df$stderr <- apply(df, 1, mf)
  out_df
}

# reference_df - subset of current_gas_cost
calculate_l2_scale <- function(df, reference_df) {
  q = "SELECT avg(estimate_marginal_ns/constant_current_gas) FROM reference_df INNER JOIN df ON df.op = reference_df.opcode"
  sqldf(q)
}

calculate_l2_row_mean <- function(df) {
  out_df = data.frame(matrix(nrow = nrow(df), ncol = 0))
  out_df$mean <- rowMeans(df, na.rm = TRUE)
  mf <- function(x){sd(df[x, ])/sqrt(length(all_envs))}
  out_df$stderr <- sapply(1:nrow(df), mf)
  out_df
}

if (params$metric == "l1") {
  calculate_scale = calculate_l1_scale
  calculate_row_mean = calculate_l1_row_mean
} else if (params$metric == "l2") {
  calculate_scale = calculate_l2_scale
  calculate_row_mean = calculate_l2_row_mean
} else {
  stop("unsupported metric")
}
```


```{r, include=FALSE}
q = c("SELECT ")
for (env in all_envs) {
  df = estimated_cost[which(estimated_cost$env==env),]
  eligible_opcodes = df[,'op'] # TODO limit opcodes for scale
  reference_df = current_gas_cost[which(current_gas_cost$base > 0),]
  scale = calculate_scale(df, reference_df)
  q <- append(q, paste0(env, ".estimate_marginal_ns/", scale, " as ", env, "_gas,"))
  q <- append(q, paste0(env, ".estimate_marginal_ns_stderr/", scale, " as ", env, "_gas_stderr,"))
}
q <- append(q, " current_gas_cost.opcode, current_gas_cost.constant_current_gas FROM current_gas_cost")
for (env in all_envs) {
  q <- append(q, paste0(" LEFT JOIN estimated_cost as ", env, " ON ", env, ".op = current_gas_cost.opcode AND ", env, ".env = '", env, "'"))
}
q <- append(q, " WHERE current_gas_cost.groups >= 0")

all_envs_cols <- paste0(all_envs, "_gas")
max_groups = max(current_gas_cost[, "groups"])

gas_schedule_comparison <- sqldf(paste(q, collapse = ""))
gas_schedule_comparison <- gas_schedule_comparison[rowSums(!is.na(gas_schedule_comparison[, all_envs_cols])) > 0, ] # remove all NA rows
```

Plot the scaled gas cost schedules.

```{r fig.width=15, fig.height=12}
for (g in 0:max_groups) {
  groups_opcode <- current_gas_cost[which(current_gas_cost$groups == g), "opcode"]
  if (length(groups_opcode) == 0) {
    next
  }
  grups_gas_schedule_comparison <- gas_schedule_comparison[which(gas_schedule_comparison$opcode %in% groups_opcode), ]
  if (nrow(grups_gas_schedule_comparison) == 0) {
    next
  }
  maximum_gas_cost = max(grups_gas_schedule_comparison[, c(all_envs_cols, "constant_current_gas")], na.rm=TRUE)
  par(mar=c(12, 4, 4, 2) + 0.1) # default is c(5, 4, 4, 2) + 0.1
  plot(grups_gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.2), xlab="", ylab="gas", las=2)
  axis(1, at=1:nrow(grups_gas_schedule_comparison), labels=grups_gas_schedule_comparison$opcode, las=2, lwd=0.1, tck=1)
  for (i in 1:length(all_envs)) {
    points(grups_gas_schedule_comparison[, paste0(all_envs[i], "_gas")], col=env_colors[i], bg=env_colors[i], xaxt='n', pch=21)
  }
  title(main="The scaled gas costs")
  legend(1, maximum_gas_cost * 1.2, c(all_envs, "Current gas cost schedule"), fill=c(env_colors[1:length(all_envs)], 'grey'), cex=1)
}
```

`r if(details){"And individual plots of scaled gas schedules for EVMs."}`

```{r fig.width=15, fig.height=12, include=details}
for (i in 1:length(all_envs)) {
  for (g in 0:max_groups) {
    groups_opcode <- current_gas_cost[which(current_gas_cost$groups == g), "opcode"]
    if (length(groups_opcode) == 0) {
      next
    }
    grups_gas_schedule_comparison <- gas_schedule_comparison[which(gas_schedule_comparison$opcode %in% groups_opcode), c(paste0(all_envs[i], "_gas"), "opcode", "constant_current_gas")]
    if (nrow(grups_gas_schedule_comparison) == 0) {
      next
    }
    maximum_gas_cost = max(grups_gas_schedule_comparison[, c(paste0(all_envs[i], "_gas"), "constant_current_gas")], na.rm=TRUE)
    par(mar=c(12, 4, 4, 2) + 0.1) # default is c(5, 4, 4, 2) + 0.1
    plot(grups_gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.2), xlab="", ylab="gas")
    axis(1, at=1:nrow(grups_gas_schedule_comparison), labels=grups_gas_schedule_comparison$opcode, las=2)
    points(grups_gas_schedule_comparison[, paste0(all_envs[i], "_gas")], col=env_colors[i], bg=env_colors[i], xaxt='n', pch=21)
    legend(1, maximum_gas_cost * 1.2, c(all_envs[i], "Current gas cost schedule"), fill=c(env_colors[i], 'grey'), cex=1)
    title(main=paste0("The scaled gas cost - ", all_envs[i]))
  }
}
```

Calculate and plot the alternative gas cost schedule.
The calculated values are:

* alternative_gas: the average of evms gas schedules, without any weights,
* alternative_gas_stderr: the standard deviation of evms gas schedules,
* alternative_gas_rel_diff: the relative difference between the current gas schedule and the alternative gas schedule, as a fraction,
* alternative_gas_rel_stderr: the relative standard deviation of evms gas schedules, relative to the alternative gas schedule, as a fraction.

```{r, include=FALSE}
if (length(all_envs_cols) == 1) { # special case
  gas_schedule_comparison$alternative_gas <- gas_schedule_comparison[, all_envs_cols]
  gas_schedule_comparison$alternative_gas_stderr <- 0
} else {
  alternative_gas = calculate_row_mean(gas_schedule_comparison[, all_envs_cols])
  gas_schedule_comparison$alternative_gas = alternative_gas$mean
  gas_schedule_comparison$alternative_gas_stderr = alternative_gas$stderr
}

gas_schedule_comparison$alternative_gas_rel_stderr <- gas_schedule_comparison$alternative_gas_stderr / gas_schedule_comparison$alternative_gas
gas_schedule_comparison$alternative_gas_rel_diff <- abs((gas_schedule_comparison$constant_current_gas - gas_schedule_comparison$alternative_gas) / gas_schedule_comparison$constant_current_gas)
```

```{r}
gas_schedule_comparison[, c('opcode', 'constant_current_gas', 'alternative_gas', 'alternative_gas_stderr', 'alternative_gas_rel_diff', 'alternative_gas_rel_stderr')]
```

```{r fig.width=15, fig.height=12}
for (g in 0:max_groups) {
  groups_opcode <- current_gas_cost[which(current_gas_cost$groups == g), "opcode"]
  if (length(groups_opcode) == 0) {
    next
  }
  grups_gas_schedule_comparison <- gas_schedule_comparison[which(gas_schedule_comparison$opcode %in% groups_opcode), c("opcode", "constant_current_gas", "alternative_gas")]
  if (nrow(grups_gas_schedule_comparison) == 0) {
    next
  }
  maximum_gas_cost = max(grups_gas_schedule_comparison[, c("constant_current_gas", "alternative_gas")], na.rm=TRUE)
  
  par(mar=c(12, 4, 4, 2) + 0.1) # default is c(5, 4, 4, 2) + 0.1
  plot(grups_gas_schedule_comparison$constant_current_gas, col='grey', xaxt='n', ylim=c(0,maximum_gas_cost * 1.2), xlab="", ylab="gas")
  axis(1, at=1:nrow(grups_gas_schedule_comparison), labels=grups_gas_schedule_comparison$opcode, las=2)
  points(grups_gas_schedule_comparison$alternative_gas, col='orange', bg='orange', xaxt='n', pch=21)
  title(main="The alternative gas cost schedule")
  legend(1, maximum_gas_cost * 1.2, c("Alternative gas cost schedule", "Current gas cost schedule"), fill=c('orange', 'grey'), cex=1)
}
```

```{r fig.width=30, fig.height=15, include=(details_arg == 100)}
# This is just for ad hoc visualization
if (details_arg == 100) {
  maximum_gas_cost = 13
  max_values = nrow(gas_schedule_comparison) - 10
  constant_current_gas_color = rgb(0.4,0.4,0.4,1.0)
  alternative_gas_color = rgb(0.1,0.2,1.0,1.0)

  plot(gas_schedule_comparison[1:max_values, 'constant_current_gas'], col=constant_current_gas_color, bg=constant_current_gas_color, xaxt='n', pch=21, ylim=c(0,maximum_gas_cost * 1.1), xlab="", ylab="")
  axis(1, at=1:max_values, labels=gas_schedule_comparison[1:max_values, 'opcode'], las=2)
  points(gas_schedule_comparison[1:max_values, 'alternative_gas'], col=alternative_gas_color, bg=alternative_gas_color, xaxt='n', pch=21, cex=1.5)
  legend(75, maximum_gas_cost, c("Alternative gas cost schedule", "Current gas cost schedule"), fill=c(alternative_gas_color, constant_current_gas_color), cex=2)
}
```

```{r fig.width=15, fig.height=10, include=(details_arg == 101)}
# BLS precompiles
if (details_arg == 101) {
  rel_ops <- c('BLS12_G1ADD', 'BLS12_G2ADD', 'BLS12_G1MSM_ARG0', 'BLS12_G2MSM_ARG0', 'BLS12_PAIRING_CHECK_ARG0', 'BLS12_PAIRING_CHECK_ARGC', 'BLS12_MAP_FP_TO_G1', 'BLS12_MAP_FP_TO_G2')
  rel_ops_tab <- data.frame(opcode=rel_ops)

  q <- "SELECT estimated_cost.op, estimated_cost.estimate_marginal_ns/current_gas_cost.constant_current_gas AS cost, estimated_cost.env
  FROM estimated_cost JOIN current_gas_cost ON estimated_cost.op = current_gas_cost.opcode
  WHERE estimated_cost.op IN rel_ops_tab"
  rel_ops_cost <- sqldf(q)

  q <- "SELECT rel_ops_cost.op, rel_ops_cost.cost/a.avg_cost AS cost, rel_ops_cost.env 
  FROM rel_ops_cost JOIN (SELECT t.env, AVG(t.cost) AS avg_cost FROM rel_ops_cost t GROUP BY t.env) a ON rel_ops_cost.env = a.env
  ORDER BY rel_ops_cost.op"
  rel_ops_scaled <- sqldf(q)

  rel_ops_tab$cost <- 1
  rel_ops_tab <- rel_ops_tab[order(rel_ops_tab$opcode),]

  max_cost = max(rel_ops_scaled$cost)
  min_cost = min(rel_ops_scaled$cost)
  par(mar=c(15, 4, 4, 2) + 0.1) # default is c(5, 4, 4, 2) + 0.1
  plot(rel_ops_tab$cost, col='grey', xaxt='n', ylim=c(min_cost * 0.9, max_cost * 1.3), xlab="", ylab="cost", las=2)
  axis(1, at=1:nrow(rel_ops_tab), labels=rel_ops_tab$opcode, las=2, lwd=0.1, tck=1)
  for (i in 1:length(all_envs)) {
    points(rel_ops_scaled[which(rel_ops_scaled$env == all_envs[i]), 'cost'], col=env_colors[i], bg=env_colors[i], xaxt='n', pch=21)
  }
  title(main="Relative computational time")
  legend(1, max_cost * 1.3, c(all_envs, "Reference value"), fill=c(env_colors[1:length(all_envs)], 'grey'), cex=1)
}
```

```{r fig.width=15, fig.height=10, include=(details_arg == 102)}
# BLS tests
if (details_arg == 102) {
  groups_for_graphs <- list( list(c(1,2), 'BLS12_G1ADD'), list(c(3,4), 'BLS12_G2ADD'), list(c(5,6), 'BLS12_G1MSM_K2'), list(c(7,8), 'BLS12_G2MSM_K2'), list(c(9,10), 'BLS12_PAIRING_CHECK'), list(c(11,12), 'BLS12_MAP_FP_TO_G1'), list(c(13,14), 'BLS12_MAP_FP_TO_G2') )
  for (groups_for_graph in groups_for_graphs) {
    rel_ops <- current_gas_cost[which(current_gas_cost$groups %in% groups_for_graph[[1]]), c('opcode', 'constant_current_gas')]
    rel_ops_tab <- rel_ops['opcode']

    q <- paste0("SELECT e1.op, e1.estimate_marginal_ns/e2.estimate_marginal_ns AS cost, e1.env 
FROM estimated_cost e1 JOIN estimated_cost e2 ON e1.env = e2.env
WHERE e1.op in rel_ops_tab AND e2.op = '", groups_for_graph[[2]], "'
ORDER BY e1.op")
    rel_ops_scaled <- sqldf(q)

    rel_ops_tab$cost <- (rel_ops[,'constant_current_gas'] / current_gas_cost[which(current_gas_cost$opcode == groups_for_graph[[2]]), 'constant_current_gas'])

    rel_ops_tab <- rel_ops_tab[order(rel_ops_tab$opcode),]

    max_cost = max(rel_ops_scaled$cost)
    min_cost = min(rel_ops_scaled$cost)
    par(mar=c(20, 4, 4, 2) + 0.1) # default is c(5, 4, 4, 2) + 0.1
    plot(rel_ops_tab$cost, col='grey', xaxt='n', pch=5, ylim=c(min_cost * 0.9, max_cost * 1.3), xlab="", ylab="cost", las=2)
    axis(1, at=1:nrow(rel_ops_tab), labels=rel_ops_tab$opcode, las=2, lwd=0.1, tck=1)
    for (i in 1:length(all_envs)) {
      points(rel_ops_scaled[which(rel_ops_scaled$env == all_envs[i]), 'cost'], col=env_colors[i], bg=env_colors[i], xaxt='n', pch=21)
    }
    title(main=paste("Relative computational time", groups_for_graph[[2]]))
    legend(1, max_cost * 1.3, c(all_envs, "EIP-2357 gas cost reference"), fill=c(env_colors[1:length(all_envs)], 'grey'), cex=1)
  }
}
```

```{r fig.width=15, fig.height=10, include=TRUE}
# (details_arg == 103)
# precompiles 0x1 - 0xa
if (details_arg == 103) {
#  rel_ops <- c('ECRECOVER', 'SHA2-256', 'RIPEMD-160', 'IDENTITY', 'MODEXP', 'ECADD', 'ECMUL', 'ECPAIRING79000', 'ECPAIRING113000', 'ECPAIRING181000', 'ECPAIRING317000', 'BLAKE2F', 'POINTEVAL')
  rel_ops <- c('SHA2-256', 'RIPEMD-160', 'IDENTITY', 'MODEXP', 'ECADD', 'ECMUL', 'ECPAIRING79000', 'ECPAIRING113000', 'ECPAIRING181000', 'ECPAIRING317000', 'BLAKE2F', 'POINTEVAL')
  rel_ops_tab <- data.frame(opcode=rel_ops)

  q <- "SELECT estimated_cost.op, estimated_cost.estimate_marginal_ns/current_gas_cost.constant_current_gas AS cost, estimated_cost.env
  FROM estimated_cost JOIN current_gas_cost ON estimated_cost.op = current_gas_cost.opcode
  WHERE estimated_cost.op IN rel_ops_tab AND estimated_cost.env != 'ethereumjs'"
  rel_ops_cost <- sqldf(q)

  q <- "SELECT rel_ops_cost.op, rel_ops_cost.cost/a.avg_cost AS cost, rel_ops_cost.env 
  FROM rel_ops_cost JOIN (SELECT t.env, AVG(t.cost) AS avg_cost FROM rel_ops_cost t GROUP BY t.env) a ON rel_ops_cost.env = a.env
  ORDER BY rel_ops_cost.op"
  rel_ops_scaled <- sqldf(q)

  rel_ops_tab$cost <- 1
  rel_ops_tab <- rel_ops_tab[order(rel_ops_tab$opcode),]

  max_cost = max(rel_ops_scaled$cost)
  min_cost = min(rel_ops_scaled$cost)
  par(mar=c(15, 4, 4, 2) + 0.1) # default is c(5, 4, 4, 2) + 0.1
  plot(rel_ops_tab$cost, col='grey', xaxt='n', ylim=c(min_cost * 0.9, max_cost * 1.5), xlab="", ylab="cost", las=2)
  axis(1, at=1:nrow(rel_ops_tab), labels=rel_ops_tab$opcode, las=2, lwd=0.1, tck=1)
  for (i in 1:length(all_envs)) {
    points(rel_ops_scaled[which(rel_ops_scaled$env == all_envs[i]), 'cost'], col=env_colors[i], bg=env_colors[i], xaxt='n', pch=21)
  }
  title(main="Relative computational time")
  legend(1, max_cost * 1.5, c(all_envs, "Reference value"), fill=c(env_colors[1:length(all_envs)], 'grey'), cex=1)
}
```

```{r, include=FALSE}
# just export the final compiled estimations to a file
# write.csv(estimated_cost, "final_estimated_cost.csv", quote=FALSE, row.names=FALSE) # estimated_cost is changed in stage4
if (params$output_comparison_file != "") {
  write.csv(gas_schedule_comparison, params$output_comparison_file, quote=FALSE, row.names=FALSE)
}
```