---
title: "R Notebook: estimation of gas costs using the measure marginal method; for Gas Cost Estimator"
output: html_notebook
---

```{r include=FALSE, fig.width=20}
library(sqldf)
library(nlme)
```


## Data preparations

Read in the `programs.csv` as generated by `src/program_generator.py` with `--fullCSV` flag on:
```{r fig.width=20}
setwd("~/sources/imapp/gas-cost-estimator/src")

program_set_codename = "exp"
programs = read.csv(paste("../../local/pg_marginal_", program_set_codename, ".csv", sep=""))
```

Unfortunately `JUMP` and `JUMPI` are currently broken in `geth` (they require a `JUMPDEST`).
Because they work in `evmone` (it prepends `JUMPDEST`) forcefully, we should drop them from the analysis.
Also `RETURNDATACOPY` program generation is broken in general (`return data out of bounds`).

```{r fig.width=20}
clean_programs = function(programs) {
  programs = programs[which(programs$opcode != "JUMP"), ]
  programs = programs[which(programs$opcode != "JUMPI"), ]
  programs = programs[which(programs$opcode != "RETURNDATACOPY"), ]
  
  # necessary to get rid of the factor levels which we just filtered out
  droplevels(programs)
}

programs = clean_programs(programs)

head(programs)
```

Read in the `result_*.csv` as generated by `src/measurements.py`.

```{r fig.width=20}
setwd("~/sources/imapp/gas-cost-estimator/src")

load_data_set <- function(env, program_set_codename, measurement_codename) {
  data_set_codename = paste(program_set_codename, "_", measurement_codename, sep="")
  filepath = paste("../../local/", env, "_pg_marginal_", data_set_codename, ".csv", sep="")
  result = read.csv(filepath)
  result$env = env
  return(result)
}
result_geth = load_data_set("geth", "exp", "total_50_4")
result_evmone = load_data_set("evmone", "exp", "total_50_4")
results = rbind(result_geth, result_evmone)
head(results)
```

Combine the two tables to see only times of the measured opcodes.

```{r fig.width=20}
measurements = sqldf("SELECT opcode, op_count, sample_id, run_id, measure_total_time_ns, env, results.program_id
                     FROM results
                     INNER JOIN
                       programs ON(results.program_id = programs.program_id)
                     ")
head(measurements)
```

### Remove outliers

Removing or not of the outliers has a large impact on the final results.
It seems that for `EXP` and `evmone` there are many outliers in the mid-low `op_count` samples.
This effectively "leverages" the estimated slope downwards, pretending `EXP`'s `evmone` to be cheaper than it really seems to be.

Switch `removed_outliers` to `FALSE` to see the comparison.

```{r}
remove_outliers <- function(df, col) {
  outliers = boxplot(df[, col] ~ df[, 'op_count'] + df[, 'env'] + df[, 'opcode'], plot=FALSE)$out
  no_outliers = df[-which(df[, col] %in% outliers), ]
  return(no_outliers)
}

removed_outliers = TRUE

if (removed_outliers) {
  measurements = remove_outliers(measurements, 'measure_total_time_ns')
}

head(measurements)
```

### Trim the initial values

```{r}
# measurements = measurements[-(which(measurements$op_count < 15)), ]
```


```{r fig.width=15}
boxplot(measure_total_time_ns ~ op_count, data=measurements[which(measurements$env == 'geth'), ], las=2, outline=removed_outliers)
boxplot(measure_total_time_ns ~ op_count, data=measurements[which(measurements$env == 'evmone'), ], las=2, outline=removed_outliers)
```

```{r}
measurements_mean = aggregate(measure_total_time_ns ~ op_count * env, measurements, mean)
  
scatter.smooth(measurements_mean[which(measurements_mean$env == 'geth'), 'op_count'], measurements_mean[which(measurements_mean$env == 'geth'), 'measure_total_time_ns'])
scatter.smooth(measurements_mean[which(measurements_mean$env == 'evmone'), 'op_count'], measurements_mean[which(measurements_mean$env == 'evmone'), 'measure_total_time_ns'])
```

```{r}
model_mean_geth = lm(measure_total_time_ns ~ op_count, data=measurements_mean[which(measurements_mean$env == 'geth'), ])
model_mean_evmone = lm(measure_total_time_ns ~ op_count, data=measurements_mean[which(measurements_mean$env == 'evmone'), ])
summary(model_mean_geth)
summary(model_mean_evmone)
```

```{r}
model_geth = lm(measure_total_time_ns ~ op_count, data=measurements[which(measurements$env == 'geth'), ])
model_evmone = lm(measure_total_time_ns ~ op_count, data=measurements[which(measurements$env == 'evmone'), ])
summary(model_geth)
summary(model_evmone)
```


```{r fig.width=15}
par(mfrow=c(4,4))
plot(model_mean_geth)
plot(model_mean_evmone)
plot(model_geth)
plot(model_evmone)
```

```{r}
compute_all <- function(program_set_codename, measurement_codename, opcode, env, removed_outliers) {
  setwd("~/sources/imapp/gas-cost-estimator/src")
  
  print(c(program_set_codename, opcode, env))
  
  programs = read.csv(paste("../../local/pg_marginal_", program_set_codename, ".csv", sep=""))
  programs = clean_programs(programs)
  results = load_data_set(env, program_set_codename, measurement_codename)
  
  measurements = sqldf(paste("SELECT opcode, op_count, sample_id, run_id, measure_total_time_ns, env, results.program_id
                       FROM results
                       INNER JOIN
                         programs ON(results.program_id = programs.program_id)
                       WHERE opcode = '", opcode, "'", sep=""))
  if (removed_outliers) {
    measurements = remove_outliers(measurements, 'measure_total_time_ns')
  }
  measurements_mean = aggregate(measure_total_time_ns ~ op_count * env, measurements, mean)
  
  par(mfrow=c(1,2))
  boxplot(measure_total_time_ns ~ op_count, data=measurements, las=2, outline=removed_outliers)
  title(main=env)
  scatter.smooth(measurements_mean[, 'op_count'], measurements_mean[, 'measure_total_time_ns'])
  title(main=opcode)
  
  model_mean = lm(measure_total_time_ns ~ op_count, data=measurements_mean)
  print(summary(model_mean))
  model = lm(measure_total_time_ns ~ op_count, data=measurements)
  print(summary(model))
  
  par(mfrow=c(2,4))
  plot(model_mean)
  plot(model)
  
  (model_mean$coefficients[['op_count']])
}
```

```{r fig.width=12}
compute_all("exp", "total_50_4", "EXP", "evmone", TRUE)
```

```{r fig.width=12}
extract_opcodes <- function(program_set_codename) {
  setwd("~/sources/imapp/gas-cost-estimator/src")
  programs = read.csv(paste("../../local/pg_marginal_", program_set_codename, ".csv", sep=""))
  programs = clean_programs(programs)
  unique(programs$opcode)
}

all_opcodes = extract_opcodes("arithmetic")
estimates = data.frame(matrix(ncol = 3, nrow = 0))
colnames(estimates) <- c('op', 'estimate_marginal_ns', 'env')
```

```{r}
for (opcode in all_opcodes) {
  estimate = compute_all("arithmetic", "total_50_4", opcode, "evmone", TRUE)
  estimates[nrow(estimates) + 1, ] = c(opcode, estimate, 'evmone')
}
```

```{r}
temp_opcodes_part1 = extract_opcodes("arithmetic_part1")
for (opcode in temp_opcodes_part1) {
  estimate = compute_all("arithmetic", "total_50_4_part1", opcode, "geth", TRUE)
  estimates[nrow(estimates) + 1, ] = c(opcode, estimate, 'geth')
}
```

```{r}
print(all_opcodes)
```

```{r}
setwd("~/sources/imapp/gas-cost-estimator/src")
write.csv(estimates, "../../local/marginal_estimated_cost.csv", quote=FALSE, row.names=FALSE)
```

